Applying State-free Reversible VAMPNets and Markov State Models to Learn Dynamics of DNA Oligonucleotides

### 2020-12-03 manuscript.tex
updates dmap section

%The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. This reveals that the mismatched C:T pairs are never bound -- a consequence of the 3spn2 excluded volume interaction -- whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, has higher values for GC-end than AT-all. Because the GC-end termini do not bind out of register, we find that they are readily able to form stabilizing contacts despite the shifted conformation of the duplex as a whole. These "shifted-loop" bonds are shown to be uniquely stable for GC-end conformations in the 5' shifted state, and their existence in the simulations is confirmed by molecular renderings sampled from these regions. Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.

@@

\subsection{Structural diversity in the F4 state informs fraying dynamics}  % shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

(might just want to move this to the SI) We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}). This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs. Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. In these conformations, one of the strands maintains some helical character while the other twists out of place, resulting in WC bonds being obstructed by the oligo backbone. These states represent another potential way in which the hybridization process (or helix reformation) can be kinetically frustrated. We observed this mode to be mostly symmetric, however there is slight tendency for the 3'T end to fray farther out of place relative to its 5'A counterpart. This might be another consequence of differential excluded volume radii in the 3spn2 force field.

>>>>

The first diffusion mode clearly delineates between the GC-end and AT-all shifted conformations and correlates highly with the average distance between the 3' end and its shifted complementary pair. As expected, this shows that mismatched C:T pairs are never bound whereas the AT-all pairs are mostly bound with occasional fraying indicated by small AT-all overlap in the GC-end region. The second diffusion mode, which correlates highly with the average distance between 3' and 5' ends, highlights structural variances within the two sequences distributions. We notice that about 40\% of GC-end configurations fall below $\Psi_3 = -0.01$  compared to only 10\% of AT-all configuration. We find that these GC-end populations share stabilizing termini contacts despite all A:T contacts being shifting out-of-register. These "shifted-loop" bonds appear uniquely stable for GC-end conformations,  and suggest an intermediate states through which hybridization is facilitated. This may account for the GC-end 5S1 $\rightarrow$ H transition being more favorable than  5S1 $\rightarrow$ D despite the state lifetime being substantially shorter than the AT-all equivalent.

%This effect may be exaggerated given that C and T base pairs are assigned slightly higher excluded volume radii in 3spn2 \citep{Hinckley2013AnHybridization}.

%Although AT-all shifted ends tend to stay bound out-of-register, the second diffusion coordinate shows some population of inert tails that fold back onto the helix.

@@

\subsection{Structural diversity in the F4 state informs fraying dynamics} 
(might just want to move this to the SI)

% shift to include this in the overall diffusion map section
%\subsubsection{Diffusion maps show ensemble of frayed states}

We applied a similar diffusion map approach to investigate the structural composition of the F4 macrostate. We build diffusion maps using 10000 frames sampled from the macrostate (figure \ref{fig:GC-core_dmaps}). Again, we were able to identify a combination of physical coordinates that closely correlated to the first two non-trivial diffusion modes. We found that a larger distance between internal A:T pairs increased the PCCA+ probability of inclusion into macrostate. This distance also correlated closely with the second diffusion map mode. Interestingly, we observed that the first non-trivial mode -- the feature that describes the most structural diversity in the system -- corresponds to difference in "overlap" distance between adjacent A or T bases and the GC core. We defined overlap as the distance between a G/C and the A/T adjacent to its native complement. At the extremes of this coordinate, one strand maintains most of its helical character while the other twists out of place. Because WC bonds are obstructed by the oligomer, the other strand is likely to fray before contacts can reform, or, less frequently, the duplex undergoes full dissociation. Interchange between these substates illustrate how the the F4 can maintain a longer lifetime compared to other A:T terminal sequences.

%This time, we set our distance metric to the same permutation-free coordinates we used to build the SRV-MSMs.
